Ricci-Ollivier Curvature of two Random Walks on the Rooted Phylogenetic Subtree-Prune-Regraft Graph

### 2015-03-31 curvature.tex
More bigger fonts; colorbar behaving.
Even bigger fonts! And I follow my own suggestions.

In this paper, we investigate curvature of the rSPR graph with respect to two random walks and connect those results to access times (also known as hitting times) for those random walks.
For this first effort we investigate random walks defined only in terms of the graph itself: the lazy uniform random walk and MCMC sampling from the uniform prior on trees.
We present a fast new algorithm for computing rSPR graphs from a set of trees, reducing the time to do so from $\OhOf{m^2n}$ to $\OhOf{mn^3}$ for a set of $m$ trees with $n$ leaves.
As the full rSPR graph contains $(2n-3)!!$ trees, this is a significant improvement in practice for exploring large subsets of the graph.
By exploiting symmetries in the rSPR graph, we were able to calculate all of the curvatures for pairs of trees with up to seven leaves.
By carefully examining the overlap in rSPR moves, we present a new method for computing the degree of a tree in the rSPR graph that allows one to select an rSPR neighbor uniformly at random in linear-time without explicitly generating the graph.
Using this method to simulate these random walks, we find that the distribution of access times between pairs of trees can be described by distance between the trees, the degrees of the trees, and the curvature.
By getting a more fine-tuned understanding of the rSPR neighborhood of pairs of vertices, we are able to give bounds on curvatures under these random walks.
In particular, we present a full characterization of the change in rSPR degree that occurs from a given rSPR move and find that even though they each count as one move, rSPR moves which move large subtrees may have greater graph effects than those that move small subtrees.
Finally, we prove that the two competing notions of Ricci-Ollivier curvature (coarse and asymptotic) are closely related when applied to the rSPR graph, differing only by a small factor inversely proportional to the rSPR degrees of the given trees.
In summary, we extend knowledge about an important graph for phylogenetics, specifically in a way that models phylogenetic MCMC search.

@@

The (coarse) Ricci-Ollivier curvature of $x$ and $y$ is then defined as:
\begin{equation}
\curvature{x, y} := 1 - \frac{W_1(m_x, m_y)}{d(x, y)}
\label{eq:curvatureDef}
\end{equation}
We use $\curvature{MH; x,y}$ to indicate curvature with respect to Metropolis-Hastings random walks (described in detail in Section~\ref{sec:random_walks}) rather than uniform random walks.
Positive curvature implies that $m_x$ and $m_y$ are ``closer'' than $x$ and $y$, zero curvature implies that they are neither closer nor farther, and negative curvature implies that $m_x$ and $m_y$ are on average ``more distant'' than $x$ and $y$.
Curvature thus provides an intuitive measure of the difficulty of moving between regions of the graph with the given random walk.

Lin et al.~\cite{lin2011ricci} defined a different notion of curvature in terms of lazy random walks which Loisel and Romon~\cite{Loisel2014-hu} dubbed the (asymptotic) Olivier-Ricci curvature.
The lazy random walk selects a neighbor according to the given distribution, but only travels to the neighbor with probability $p$.
We denote the coarse curvature of the lazy random walk between two trees $x$ and $y$ by $\curvature[p]{x,y}$.
For example, $\curvature[\frac{1}{4}]{x,y}$ describes the curvature of the lazy random walk that follows the given distribution with probability $\frac{1}{4}$ and remains stationary with probability $\frac{3}{4}$.
For the lazy uniform random walk, the mass assignment $m^p_x$ is $x \cup N(x)$, with each neighbor assigned mass $\frac{p}{\degree{x}}$ and $x$ assigned the remaining $1 - p$ mass.
The (asymptotic) Ricci-Ollivier curvature of $x$ and $y$ is then:
\begin{equation}
\ric{x, y} := \lim_{p \rightarrow 0} \frac{\curvature[p]{x,y}}{p}
\label{eq:curvatureDef}
\end{equation}
This definition of curvature is invariant of $p$ for small enough $p$~\cite{Loisel2014-hu} and can be used to avoid parity problems on graphs where the uniform random walk is periodic without choosing a specific laziness parameter (e.g. Ollivier often considered $\curvature[\frac{1}{2}]{x,y}$ for this purpose).
As we prove in Lemma~\ref{lem:asymptotic} the two notions of rough and asymptotic curvature differ only by a small factor bounded by $\frac{2}{\max(\degree{x}, \degree{y})}$ between adjacent trees and are equal for nonadjacent trees.
We define $\ric{MH; x, y}$ analogously to $\curvature{MH; x, y}$.

@@

\section{Conclusion and future work}
In summary, we have learned characteristics of the phylogenetic rSPR graph underlying phylogenetic inference substantially beyond previous work concerning graph diameter and vertex degree.
We did so by developing theoretical and computational frameworks to bound and compute Ricci-Ollivier curvature of the rSPR graph.
We find that curvature, along with degree and distance, determine the early dynamics of hitting times for random walks.
Moreover, we proved that rSPR graph degree changes depend quadratically on the product of the size of the regrafted subtree with its change in depth as well as that rSPR graph tends toward flatness with respect to rSPR moves that move asymptotically small subtrees.
Finally, we proved that the two competing definitions of Ollivier-Ricci curvature, coarse and asymptotic, are closely related with respect to uniform and Metropolis-Hastings walks on the rSPR graph.

>>>>

In this paper, we investigate curvature of the rSPR graph with respect to two random walks and connect those results to access times (also known as hitting times) for those random walks.
For this first effort we investigate random walks defined only in terms of the graph itself: the lazy uniform random walk and MCMC sampling from the uniform prior on trees.
We present a fast new algorithm for computing rSPR graphs from a set of trees, reducing the time to do so from $\OhOf{m^2n}$ to $\OhOf{mn^3}$ for a set of $m$ trees with $n$ leaves.
As the full rSPR graph contains $(2n-3)!!$ trees, this is a significant improvement in practice for exploring large subsets of the graph.
By exploiting symmetries in the rSPR graph, we were able to calculate all of the curvatures for pairs of trees with up to seven leaves.
By carefully examining the overlap in rSPR moves, we present a new method for computing the degree of a tree in the rSPR graph that allows one to select an rSPR neighbor uniformly at random in linear-time without explicitly generating the graph.
Using this method to simulate these random walks, we find that the distribution of access times between pairs of trees can be described by distance between the trees, the degrees of the trees, and the curvature.
By getting a more fine-tuned understanding of the rSPR neighborhood of pairs of vertices, we are able to give bounds on curvatures under these random walks.
In particular, we present a full characterization of the change in rSPR degree that occurs from a given rSPR move and find that even though they each count as one move, rSPR moves which move large subtrees may have greater graph effects than those that move small subtrees.
Finally, we prove that the two related notions of Ricci-Ollivier curvature (coarse and asymptotic) are close, and in fact identical in the limit of many taxa, when applied to the rSPR graph.
In summary, we extend knowledge about an important graph for phylogenetics, specifically in a way that models phylogenetic MCMC search.

The automated computational analysis code can be found at \url{https://github.com/matsengrp/curvature}.

@@

The \emph{coarse} Ricci-Ollivier curvature of $x$ and $y$ is then defined as:
\begin{equation}
\curvature{x, y} := 1 - \frac{W_1(m_x, m_y)}{d(x, y)}
\label{eq:curvatureDef}
\end{equation}
We use $\curvature{MH; x,y}$ to indicate curvature with respect to Metropolis-Hastings random walks (described in detail in Section~\ref{sec:random_walks}) rather than uniform random walks.
Positive curvature implies that $m_x$ and $m_y$ are ``closer'' than $x$ and $y$, zero curvature implies that they are neither closer nor farther, and negative curvature implies that $m_x$ and $m_y$ are on average ``more distant'' than $x$ and $y$.
Curvature thus provides an intuitive measure of the difficulty of moving between regions of the graph with the given random walk.

Lin et al.~\cite{lin2011ricci} defined a different notion of curvature in terms of lazy random walks which Loisel and Romon~\cite{Loisel2014-hu} dubbed the \emph{asymptotic} Olivier-Ricci curvature.
The lazy random walk selects a neighbor according to the given distribution, but only travels to the neighbor with probability $p$.
We denote the coarse curvature of the lazy random walk between two vertices $x$ and $y$ by $\curvature[p]{x,y}$.
For example, $\curvature[\frac{1}{4}]{x,y}$ describes the curvature of the lazy random walk that follows the given distribution with probability $\frac{1}{4}$ and remains stationary with probability $\frac{3}{4}$.
For the lazy uniform random walk, the mass assignment $m^p_x$ is $x \cup N(x)$, with each neighbor assigned mass $\frac{p}{\degree{x}}$ and $x$ assigned the remaining $1 - p$ mass.
The (asymptotic) Ricci-Ollivier curvature of $x$ and $y$ is then:
\begin{equation}
\ric{x, y} := \lim_{p \rightarrow 0} \frac{\curvature[p]{x,y}}{p}
\label{eq:curvatureDef}
\end{equation}
This definition of curvature is invariant of $p$ for small enough $p$~\cite{Loisel2014-hu} and can be used to avoid parity problems on graphs where the uniform random walk is periodic without choosing a specific laziness parameter (e.g. Ollivier often considered $\curvature[\frac{1}{2}]{x,y}$ for this purpose).
As we prove in Lemma~\ref{lem:asymptotic} the two notions of rough and asymptotic curvature differ only by a small factor bounded by $\frac{2}{\max(\degree{x}, \degree{y})}$ between adjacent vertices and are equal for nonadjacent vertices.
We define $\ric{MH; x, y}$ analogously to $\curvature{MH; x, y}$.

@@

\section{Conclusion and future work}
In summary, we have learned characteristics of the phylogenetic rSPR graph underlying phylogenetic inference substantially beyond previous work concerning graph diameter and vertex degree.
We did so by developing theoretical and computational frameworks to bound and compute Ricci-Ollivier curvature of the rSPR graph.
We find that curvature, along with degree and distance, determine the early dynamics of hitting times for random walks.
Moreover, we proved that rSPR graph degree changes depend quadratically on the product of the size of the regrafted subtree with its change in depth as well as that rSPR graph tends toward flatness with respect to rSPR moves that move asymptotically small subtrees.
Finally, we proved that the coarse and asymptotic definitions of Ollivier-Ricci curvature, coarse and asymptotic, are closely related with respect to uniform and Metropolis-Hastings walks on the rSPR graph.
